---
output: github_document
---
  
<!-- README.md is generated from README.Rmd. Please edit that file -->
  
[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
[![Bioconductor devel build status](http://bioconductor.org/shields/build/devel/bioc/GenomicTuples.svg)](http://bioconductor.org/checkResults/devel/bioc-LATEST/GenomicTuples/)
[![Linux Build Status](https://travis-ci.org/PeteHaitch/GenomicTuples.svg?branch=master)](https://travis-ci.org/PeteHaitch/GenomicTuples)
[![Bioconductor downloads](http://bioconductor.org/shields/downloads/GenomicTuples.svg)](http://bioconductor.org/packages/stats/bioc/GenomicTuples.html)
[![Coverage Status](https://img.shields.io/codecov/c/github/PeteHaitch/GenomicTuples/master.svg)](https://codecov.io/github/PeteHaitch/GenomicTuples?branch=master)
[![JOSS](http://joss.theoj.org/papers/10.21105/joss.00020/status.svg)](http://dx.doi.org/10.21105/joss.00020)
[![DOI](https://zenodo.org/badge/22085/PeteHaitch/GenomicTuples.svg)](https://zenodo.org/badge/latestdoi/22085/PeteHaitch/GenomicTuples)

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

R package: GenomicTuples
---

__GenomicTuples__ is an R/Bioconductor package that defines general purpose containers for storing and manipulating _genomic tuples_. A genomic tuple of size `m` is of the form `chromosome:strand:{pos_1, pos_2, ..., pos_m}` where `pos_1` < `pos_2` < `...` < `pos_m` are positions along the chromosome. __GenomicTuples__ aims to provide functionality for tuples of genomic co-ordinates that are analogous to those available for genomic ranges in the [__GenomicRanges__](http://bioconductor.org/packages/GenomicRanges/) R/Bioconductor package.

## Installation

Most users will want to install __GenomicTuples__ using the current release of Bioconductor using:

```{r, eval = FALSE}
## try http:// if https:// URLs are not supported
source("https://bioconductor.org/biocLite.R")
biocLite("GenomicTuples")
```

The master branch of this repository is the development version of the package. The development version of __GenomicTuples__ can only be installed using the development version of Bioconductor. Please first read these [instructions on installing the development version of Bioconductor](http://www.bioconductor.org/developers/how-to/useDevel/); __GenomicTuples__ can then be installed by:

```r
source("http://bioconductor.org/biocLite.R")
useDevel()
biocLite("GenomicTuples")
```

## Quick demo

Here we use the __GenomicTuples__ package to define two _GTuples_ objects, one containing 5 3-tuples and one containing 3 3-tuples, demonstrate how to identify 'equal' genomic tuples, and how this calculation would be incorrect if we were to mistakenly treat these genomic tuples as genomic ranges:

```{r, eval = TRUE, echo = TRUE, message = FALSE}
# Load the package
library(GenomicTuples)

# Create a GTuples object containing 5 3-tuples
x <- GTuples(seqnames = c('chr1', 'chr1', 'chr1', 'chr1', 'chr2'), 
               tuples = matrix(c(10L, 10L, 10L, 10L, 10L, 20L, 20L, 20L, 25L, 
                                 20L, 30L, 30L, 35L, 30L, 30L), ncol = 3), 
               strand = c('+', '-', '*', '+', '+'))
x
# Create a new GTuples object containing 3 3-tuples by subsetting x
y <- x[2:4]
y

# Find equal genomic tuples
subsetByOverlaps(x, y, type = "equal")
# Find equal genomic tuples when ignoring strand
subsetByOverlaps(x, y, type = "equal", ignore.strand = TRUE)

# Note that if were to mistakenly treat these as genomic ranges, then the set 
# of 'equal' genomic tuples would be incorrect since treating these tuples as 
# ranges ignores the "internal positions" (pos2).
# The set of overlaps when correctly treated as genomic tuples:
findOverlaps(x, y, type = "equal")
# The set of overlaps when incorrected treated as genomic ranges:
findOverlaps(as(x, "GRanges"), as(y, "GRanges"), type = 'equal')
```

__GenomicTuples__ includes extensive documentation available through the R help system:

```{r, eval = FALSE}
# See all documentation for the package
help(package = "GenomicTuples")
# See documentation for the GTuples class
?GTuples
```

The package also includes a comprehensive vignette that explains in greater detail the different between a genomic tuple and a genomic range, when genomic tuples may be useful, and common operations on genomic tuples. The vignette can be viewed at [http://bioconductor.org/packages/release/bioc/vignettes/GenomicTuples/inst/doc/GenomicTuplesIntroduction.html](http://bioconductor.org/packages/release/bioc/vignettes/GenomicTuples/inst/doc/GenomicTuplesIntroduction.html) or accessed from R using:

```{r, eval = FALSE}
vignette("GenomicTuplesIntroduction", package = "GenomicTuples")
```

## Need help or want to get involved?

I recommend that questions seeking support in using the software are posted to the Bioconductor support forum - [https://support.bioconductor.org/](https://support.bioconductor.org/) - where they will attract not only my attention but that of the wider Bioconductor community.

Code contributions, bug reports (and fixes!), and feature requests are most welcome. Please make any pull requests against the master branch at [https://github.com/PeteHaitch/GenomicTuples](https://github.com/PeteHaitch/GenomicTuples) and file issues at [https://github.com/PeteHaitch/GenomicTuples/issues](https://github.com/PeteHaitch/GenomicTuples/issues)

## Citation

__GenomicTuples__ has been published in The Journal of Open Source Software, [http://joss.theoj.org/papers/10.21105/joss.00020](http://joss.theoj.org/papers/10.21105/joss.00020).

If you use __GenomicTuples__, please cite the paper and software version. This can be done as follows: 

```{r, citation}
citation("GenomicTuples")
packageVersion("GenomicTuples")
```

## License

__GenomicTuples__ is licensed under [Artistic 2.0](https://www.r-project.org/Licenses/Artistic-2.0).
